### 2015-02-04 Studies/2012_PG_County/Report/PG_Co_Fire.tex
PG Co: update report and figures based on initial reviewer comments.

\chapter*{\centering Abstract}
Fire Dynamics Simulator (FDS), which is a fire model that is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on February 24, 2012, within a single-story, single-family residential structure in Riverdale Heights, MD, that resulted in the serious injury of two firefighters. The inputs for the FDS simulation are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement and resulted in ventilation limited (fuel rich) fire conditions in the basement area. After the front door was opened the interior stairwell, aided by strong winds blowing into the basement, acted as a chimney for hot gases in the basement to flow towards regions of lower pressure, out through the front door. The temperature of the gases in the interior stairwell was estimated to be in excess of 400~$^{\circ}$C (750~$^{\circ}$F). Two firefighters were located in the flow path between the interior stairwell door and the front door of the structure, were exposed to these elevated temperatures, and suffered serious burn injures. Five additional firefighters were injured during fire operations on the first floor.
\pagenumbering{gobble}

@@

On February 24, 2012, a fire in residential structure injured seven Prince George's County firefighters. NIST examined the fire dynamics of this incident at the request of the Prince George's Country Fire/Emergency Medical Services (EMS) Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this single-story plus basement residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Prince George's Fire Department likely experienced during the course of their interior operations.
\item To provide a basis for discussing tactical operations.
\end{enumerate}

@@

Truck 809 Forcible Entry was too large in physical stature to fit through the small window opening that he found on [the front of the structure]. Truck 809 Forcible Entry was unable to self-evacuate, and remained trapped inside. Due to rapidly deteriorating conditions, Truck 809 Officer was forced to self-rescue through another small window on [the front of the structure].

After exiting the structure, Truck 809 Officer screamed (to those on the exterior) that a firefighter was trapped inside. Truck 809 Officer then proceeded to the front door \ldots in an attempt to search for and rescue Truck 809 Forcible Entry, who was trapped inside. At this point multiple firefighters on the exterior assisted Truck 809 Officer in the search, rescue, and removal of Truck 809 Forcible Entry. Both Truck 809 Officer and Truck 809 Forcible Entry sustained serious injuries and were transported to MedStar/Washington Hospital Center.

While the interior operations were occurring, Truck 809 Driver and Truck 809 Hook/Ladders were on the exterior throwing ladders to the structure. Truck 809 Hook/Ladders threw two ladders [front and left side], while Truck 809 Driver threw one ladder ([front side]). 
\end{quote}  
From a combination of on-scene video and the narrative accounts from the Prince George's County Fire/EMS Department report~\cite{PGCounty2013}, Table~\ref{tab:fire_info} shows an overview of the timeline of events. Table~\ref{tab:fire_info} represents the best estimate of the events and therefore all times should be considered as approximate.

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a one story, single-family structure with a basement. The structure was platform-framed wood construction with a gable roof. \footnote{Platform-framed construction means that floor joists are completely covered with a sub-floor. This forms a platform for which walls are built from.} The structure had a first floor footprint of 9.14~m (30~ft) by 7.77~m (25.5~ft) with the basement being slightly smaller, 9.14~m (30~ft) by 7.16~m (23.5~ft), as show in Fig.~\ref{fig:simp_geom}. This resulted in a first floor area of 71~m$^2$ (764~ft$^2$) and a basement floor area of 65~m$^2$ (700~ft$^2$). The interior walls were composed of gypsum board over insulated wood framing~\cite{PGCounty2013}. There was a single flight of stairs in the center of the structure which connected the first floor to the basement.

@@

The fire, which is described in Section~\ref{fire}, started in the basement and spread to the stairs which connect to the first floor. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~\ref{fig:exterior} shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the basement of the structure. Fully dimensioned drawings of the interior of the basement and first floor are shown in Figs.~\ref{fig:basement} and \ref{fig:first_floor}, respectively.

@@

\section{Fire}
\label{fire}
To estimate the type of fuel, fire size, and structure ventilation, on-scene videos and post-incident reports were used. Based on these sources of information, the FDS model includes two source fire locations in the basement: one fire in the kitchen area and one in the adjacent bathroom (Fig.~\ref{fig:basement_fire}). At the time of the fire, the structure was vacant so the the dominant fuel was wood: wood paneling on walls, exposed ceiling joists, sub-floor, cabinets, and stairs.

@@

To calculate the total fire size, the appropriate burning area of the basement needed to be determined. Based on the floor plan of the structure and post-incident images, two distinct fires were considered: 1 source in the basement bathroom and 1 source in the basement kitchen. The bathroom fire was estimated to be comprised of wood paneling along the four walls (17.3~m$^2$ or 186~ft$^2$) and wood subfloors and studs along the ceiling (2.9~m$^2$ or 31~ft$^2$). This resulted in 20.2~m$^2$ (217~ft$^2$) of wood burning surface area or a 1~MW fire using a 50~kW/m$^2$ HRRPUA. The remainder of the wood combustable load in the basement was lumped into the localized kitchen fire. The ceiling and wall areas included in the basement kitchen and main living area totaled approximately 86~m$^2$ (926~ft$^2$). This resulted in a 4.3~MW fire.

@@

A comparison of the differences between the HRR inside the structure (Fig.~\ref{fig:PG_Total_HRR}) and the total HRR (Fig.~\ref{fig:PG_9MW_HRR}) over the first 100 seconds of the simulation indicated that on average there was the equivalent of an approximate 0.675 MW fire(s) occurring through open external vents (basement windows). This aligns with observed post-fire damage and post-incident report~\cite{PGCounty2013}. Fig.~\ref{fig:ext_fire_dam} shows a post-fire image of the back left corner of the structure as well as a Smokeview rendering of the fire from the FDS simulation prior to the front door being opening by firefighters.

@@

\begin{figure}[!ht]
\includegraphics[width=5in]{../Figures/PG_First_9MW_HRR}
\caption[Prescribed and first floor calculated HRR vs. time from the simulation.]
{Comparison of total prescribed and first floor calculated HRRs from the simulation. The vertical lines indicate the times at which ventilation occurred.}
\label{fig:PG_First_9MW_HRR}
\end{figure}

@@

The report from Prince George's County Fire/EMS~\cite{PGCounty2013} indicated that two firefighters from Truck 809 were injured after making entry to the first floor and had to either bail out through a window or be rescued through the front door. From Section~\ref{temp}, the conditions on the first floor were initially tenable. After the front door was opened, however, the simulation results indicated that a high hazard area existed in the first floor that exceeded the conditions of a Class~III exposure (temperatures greater than 260~$^{\circ}$C or 500~$^{\circ}$F)~\cite{Donnelly2006}.

@@

\section{Assessing the Hazard}
\label{assessing_hazard}
A person is susceptible to second-degree burn injuries if exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate material in the facepiece of a self-contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F)~\cite{nfpa2013standard}. Prolonged exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him or her at risk. Exposure of equipment to temperatures of 260~$^{\circ}$C (500~$^{\circ}$F) represents a Class~III exposure~\cite{Donnelly2006}. Firefighters are at increased risk levels when encountering Class~III exposure conditions for more than 5~minutes~\cite{Donnelly2006}.

Figures~\ref{fig:int_temp_99s} to \ref{fig:int_temp_142s} show two interior views of temperature contours at particular snapshots in time. The top image is a planar view through the middle of the front door (from the front to the back of the house) which shows the conditions in the living area on the first floor and the conditions in the basement just past the bottom of the stairs. The bottom image shows temperature contours through the middle of the interior stairwell which connected the basement to the first floor of the structure (looking from the rear of the structure towards the front). Prior to the front door being opened, (Fig.~\ref{fig:int_temp_99s} shows that temperatures in excess of 200~$^{\circ}$C (400~$^{\circ}$F) were confined to basement of the structure: specifically the rear of the structure and under the stairwell. One second after the door was opened, hazard in the structure increased within the structure, the most drastic change occurring at the top of the stairwell as peak simulated temperatures exceeded 400~$^{\circ}$C (750~$^{\circ}$F). At the approximate time firefighters made entry to the first floor, Fig.~\ref{fig:int_temp_142s}, the conditions within the flowpath (interior stairwell through front door) had become more severe with temperatures in excess of 400~$^{\circ}$C (750~$^{\circ}$F), which exceeds the Class III exposure of 260~$^{\circ}$C (500~$^{\circ}$F). These high-hazard conditions within the flowpath remained until the modeled HRR was reduced to simulate the action of applying water the fire.

@@

In this fire incident, the initial failure of the basement windows with strong incident winds on the rear side of the structure coincided with a rapid change in the thermal conditions in the interior stairwell and first floor after the front door to the structure was opened. The open front door resulted in the establishment of a flow path within the interior stairwell with highly hazardous conditions. These conditions would be equivalent to or greater than a Class~III exposure~\cite{Donnelly2006}.

@@

Fire suppression efforts should be coordinated with interior operations and ventilation procedures to reduce thermal hazards related to flow paths within a structure. Ongoing research by NIST, Underwriters Laboratories (UL), and others has demonstrated that applying water from the exterior into the fire area of a structure (prior to the start of interior operations) can significantly improve the safety of firefighters by reducing the thermal hazard from the fire and reducing the potential for developing high velocity hot gas flows within the structure~\cite{madrzykowski2009fire, kerber2009fire}.

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards regions of lower pressure through the open front door of the structure. After the front door was opened, a flow path was established between the basement and the front side of the structure (the front door). The opening of the front door resulted in a rapid change in the conditions within the flow path. In the interior stairwell, the flow velocities were greater than 9~m/s (20~mph) and the temperature of the gases was estimated to be in excess of 400~$^{\circ}$C (750~$^{\circ}$F), which exceeds the Class III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F).

@@

The authors would like to thank XXXXX of the Prince George's County Fire/EMS Department for their assistance with this study and their dedication to improving firefighter safety. The authors would also like to thank John Culbertson, Director of the Montana Fire Services Training School, for his assistance in researching and compiling the flow path related LODD/LODI incidents shown in Table~\ref{tab:LODD}. Finally, the authors would like to thank Joseph Willi of the University of Illinois Urbana-Champaign for his assistance in creating dimensioned drawings of the structure.

>>>>

\chapter*{\centering Abstract}
Fire Dynamics Simulator (FDS), which is a fire model that is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on February 24, 2012, within a single-story, single-family residential structure in Riverdale Heights, MD, that resulted in the serious injury of two Price George's County firefighters. The inputs for the FDS simulation are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement which generated ventilation limited (fuel rich) fire conditions in the basement area. After the front door was opened, the interior stairwell, aided by strong winds blowing into the basement, acted as a chimney for hot gases in the basement to flow towards regions of lower pressure, out through the front door. The temperature of the gases in the interior stairwell was estimated to be in excess of 400~$^{\circ}$C (750~$^{\circ}$F). Two firefighters were located in the flow path between the interior stairwell door and the front door of the structure, were exposed to these elevated temperatures, and suffered serious burn injures. Five additional firefighters were injured during fire operations on the first floor.
\pagenumbering{gobble}

@@

On February 24, 2012, a fire in residential structure injured seven Prince George's County firefighters. NIST examined the fire dynamics of this incident at the request of the Prince George's Country Fire/Emergency Medical Services (EMS) Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this single-story residential structure with a basement using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Prince George's Fire Department likely experienced during the course of their interior operations.
\item To provide a basis for discussing tactical operations.
\end{enumerate}

@@

Truck 809 Forcible Entry was unable to self-evacuate, and remained trapped inside. Due to rapidly deteriorating conditions, Truck 809 Officer was forced to self-rescue through another small window on [the front of the structure].

After exiting the structure, Truck 809 Officer [alerted other firefighters] that a firefighter was trapped inside. Truck 809 Officer then proceeded to the front door \ldots in an attempt to search for and rescue Truck 809 Forcible Entry, who was trapped inside. At this point multiple firefighters on the exterior assisted Truck 809 Officer in the search, rescue, and removal of Truck 809 Forcible Entry. Both Truck 809 Officer and Truck 809 Forcible Entry sustained serious injuries and were transported to MedStar/Washington Hospital Center.

\end{quote}  
From a combination of on-scene video and the narrative accounts from the Prince George's County Fire/EMS Department report~\cite{PGCounty2013}, Table~\ref{tab:fire_info} shows an overview of the timeline of events. Table~\ref{tab:fire_info} represents the best estimate of the events.

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a one story, single-family structure with a basement. The structure was platform-framed wood construction with a gable roof. \footnote{Platform-framed construction means that floor joists are completely covered with a sub-floor. This forms a platform for which walls are built from.} The structure had a first floor footprint of 9.14~m (30~ft) by 7.77~m (25.5~ft) with the basement being slightly smaller, 9.14~m (30~ft) by 7.16~m (23.5~ft), as show in Fig.~\ref{fig:simp_geom}. This resulted in a first floor area of 71~m$^2$ (764~ft$^2$) and a basement floor area of 65~m$^2$ (700~ft$^2$). The interior walls were composed of gypsum board over wood framing~\cite{PGCounty2013}. There was a single flight of stairs in the center of the structure which connected the first floor to the basement.

@@

The fire, which is described in Section~\ref{fire}, started in the basement and spread to the stairs which connect to the first floor. The door at the top of the stairs was open at the time of the fire. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~\ref{fig:exterior} shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the basement of the structure. Fully dimensioned drawings of the interior of the basement and first floor are shown in Figs.~\ref{fig:basement} and \ref{fig:first_floor}, respectively.

@@

\section{Fire}
\label{fire}
To estimate the type of fuel, fire size, and structure ventilation, on-scene videos and post-incident reports were used. Based on these sources of information, the FDS model includes two source fire locations in the basement: one fire in the kitchen area and one in the adjacent bathroom (Fig.~\ref{fig:basement_fire}). At the time of the fire, the structure was unfurnished so the the dominant fuel was wood: wood paneling on walls, exposed ceiling joists, sub-floor, cabinets, and stairs.

@@

To calculate the total fire size, the appropriate burning area of the basement needed to be determined. Based on the floor plan of the structure and post-incident images, two distinct fires were considered: 1 source in the basement bathroom and 1 source in the basement kitchen. The bathroom fire was estimated to be comprised of wood paneling along the four walls (17.3~m$^2$ or 186~ft$^2$) and wood subfloors and studs along the ceiling (2.9~m$^2$ or 31~ft$^2$). This resulted in 20.2~m$^2$ (217~ft$^2$) of wood burning surface area or a 1~MW fire using a 50~kW/m$^2$ HRRPUA. The remainder of the wood combustable load in the basement was lumped into the localized kitchen fire. The ceiling and wall areas included in the basement kitchen, stairway, and main living area totaled approximately 86~m$^2$ (926~ft$^2$). This resulted in a 4.3~MW fire.

@@

A comparison of the differences between the HRR inside the structure (Fig.~\ref{fig:PG_Total_HRR}) and the total HRR (Fig.~\ref{fig:PG_9MW_HRR}) over the first 100 seconds of the simulation indicated that on average there was the equivalent of an approximate 0.7 MW fire(s) occurring through open external vents (basement windows). This aligns with observed post-fire damage and post-incident report~\cite{PGCounty2013}. Fig.~\ref{fig:ext_fire_dam} shows a post-fire image of the back left corner of the structure as well as a Smokeview rendering of the fire from the FDS simulation prior to the front door being opening by firefighters.

@@

\begin{figure}[!ht]
\includegraphics[width=5in]{../Figures/PG_First_9MW_HRR}
\caption[First floor calculated HRR vs. time from the simulation.]
{First floor calculated HRR from the simulation. The vertical lines indicate the times at which ventilation occurred.}
\label{fig:PG_First_9MW_HRR}
\end{figure}

@@

The report from Prince George's County Fire/EMS~\cite{PGCounty2013} indicated that two firefighters from Truck 809 were injured after making entry to the first floor and had to either bail out through a window or be rescued through the front door. From Section~\ref{temp}, the conditions on the first floor were initially tenable. After the front door was opened, however, the simulation results indicated that a high hazard area existed in the first floor that exceeded the conditions that would put a firefighter at risk for burn injuries (temperatures greater than 260~$^{\circ}$C or 500~$^{\circ}$F)~\cite{nfpa2013standard}.

@@

\section{Assessing the Hazard}
\label{assessing_hazard}
A person is susceptible to second-degree burn injuries if skin temperatures exceed 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate material in the facepiece of a self-contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures up to 260~$^{\circ}$C (500~$^{\circ}$F) or 10~kW/m$^2$~\cite{nfpa2013standard}. Exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him or her at risk.

Figures~\ref{fig:int_temp_99s} to \ref{fig:int_temp_142s} show two interior views of temperature contours at particular snapshots in time. The top image is a planar view through the middle of the front door (from the front to the back of the house) which shows the conditions in the living area on the first floor and the conditions in the basement just past the bottom of the stairs. The bottom image shows temperature contours through the middle of the interior stairwell which connected the basement to the first floor of the structure (looking from the rear of the structure towards the front). Prior to the front door being opened, (Fig.~\ref{fig:int_temp_99s} shows that temperatures in excess of 200~$^{\circ}$C (400~$^{\circ}$F) were confined to basement of the structure: specifically the rear of the structure and under the stairwell. One second after the door was opened, hazard in the structure increased within the structure, the most drastic change occurring at the top of the stairwell as peak simulated temperatures exceeded 400~$^{\circ}$C (750~$^{\circ}$F). At the approximate time firefighters made entry to the first floor, Fig.~\ref{fig:int_temp_142s}, the conditions within the flowpath (interior stairwell through front door) had become more severe with temperatures in excess of 400~$^{\circ}$C (750~$^{\circ}$F), where a fully protected firefighter would be susceptible to burn injuries. These high-hazard conditions within the flowpath remained until the modeled HRR was reduced to simulate the action of applying water the fire.

@@

In this fire incident, the initial failure of the basement windows with strong incident winds on the rear side of the structure coincided with a rapid change in the thermal conditions in the interior stairwell and first floor after the front door to the structure was opened. The open front door resulted in the establishment of a flow path within the interior stairwell from the high pressure side of the structure to the low presure side with highly hazardous conditions.

@@

First suppression efforts should take place at the same level as the first with the wind at the back of firefighters. Ongoing research by NIST, Underwriters Laboratories (UL), and others has demonstrated that applying water from the exterior into the fire area of a structure (prior to the start of interior operations) can significantly improve the safety of firefighters by reducing the thermal hazard from the fire and reducing the potential for developing high velocity hot gas flows within the structure~\cite{madrzykowski2009fire, kerber2009fire}.

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards regions of lower pressure through the open front door of the structure. After the front door was opened, a flow path was established between the basement and the front side of the structure (the front door). The opening of the front door resulted in a rapid change in the conditions within the flow path. In the interior stairwell, the flow velocities were greater than 9~m/s (20~mph) and the temperature of the gases was estimated to be in excess of 400~$^{\circ}$C (750~$^{\circ}$F), which puts fully protected firefighters at risk for burn injuries.

@@

The authors would like to thank Chief Marc Bashoor of the Prince George's County Fire/EMS Department for their assistance with this study and their dedication to improving firefighter safety. The authors would like to thank members of the Safety Investigation Team for compiling the report: Deputy Chief Scott Hoglander, Battalion Chief Sayshan Conver-White, Chief Richard Bowers, Deputy Chief William Goldfeder, Assistant Chief Michael Nelson, Volunteer Safety Division Chief Stefan Ganset, Paramedic Captain Roland Berg, Lieutenant Philip Bird, Lieutenant Daniel Schrader, Investigator Aaron Tyler, Fire Protection Engineer Lee McCarthy, President Charles Walker, and Administrative Aide Xiomara Lozano-Chevez. The authors would also like to thank John Culbertson, Director of the Montana Fire Services Training School, for his assistance in researching and compiling the flow path related LODD/LODI incidents shown in Table~\ref{tab:LODD}. Finally, the authors would like to thank Joseph Willi of the University of Illinois Urbana-Champaign for his assistance in creating dimensioned drawings of the structure.
